An Efficient Feedback Calibration Algorithm for Direct Imaging Radio Telescopes

### 2016-06-22 cal_demo/ms.tex
Made references to figures consistent

We show schematically the process of calibrating a direct imaging correlator in 
figure~\ref{fig:schematic}. Computationally expensive steps that must be performed in near real-time 
are shown inside the gray box. The uncalibrated antenna signals are tapped out after the 
F-engine and correlated against the output image pixel of interest (i.e., equation~\ref{eq:Cna} is 
computed in near real-time). The correlated values are then used to estimate the gains
using equation~\ref{eq:cal_solution}, and additional fitting if desired. This step does not have
the speed requirement of the steps in the gray box because the correlations have been 
averaged in time.
The gains are then passed back 
to the correlator to update the calibration for subsequent integration intervals.

@@

The calibration loop continues by updating the gain estimates every \caliter~time samples. The 
phases of our gain estimates are shown in figure~\ref{fig:sim_phase} for 20 such iterations. 
The phase error plotted is the phase relative to the true gain for each antenna (various colored 
lines). One antenna was used as a reference to fix the absolute phase, so has zero phase 
error. The other 50 antennas are shown to have error spanning $2\pi$ initially, and after about 
10 iterations lock into a solution, settling down to noise levels around iteration 12 (0.12~s). We 
stop the simulation when the updated gains trace the thermal noise of the simulated sources, 
which can be seen by the coherence of the 50 antenna gains after iteration 12.

@@

The estimated gain amplitudes for the simulations are shown in figure~\ref{fig:sim_amp}. The 
quantity plotted is the magnitude of the estimated gains over the true gains, $\left|g^{(n)}_a/
g_a\right|$, which places all antennas on the same scale. We can see the amplitudes 
converge toward their true values around the same time as the phases (iteration $\sim$12). At 
the beginning of calibration we can see the importance of the damping factor. At $n=0$, a couple of 
gains are shown to have abnormally high amplitude estimates, notably one about 3.3 times its 
true value (red line). These unbalanced high estimates caused the entire set of gains to be 
under estimated at $n=1$, even with a damping factor of 0.35. By $n=5$ the unbalanced 
amplitudes have been damped out and the calibration continues. Without the damping factor, 
the oscillation seen in the first couple iterations would have been significantly larger and taken 
much longer to fade out.

@@

Images created at the beginning of calibration and at the end are shown in 
figure~\ref{fig:sim_images}. Each image is obtained over 400 time sample integration, corresponding to 
all snapshot images created with a given set of gain estimates. The left panel shows the image 
produced with our initialized unity gains. Because the phases are completely random, the 
image is essentially noise with the primary beam evident. After 20 calibration iterations, the image is far 
more clear, shown in the middle panel. 
The ten sources are visible, along with rumble throughout the image. By comparing with
a perfectly calibrated image (right panel), we see that the rumble and the shapes of the
point sources are matched between the two images. We therefore conclude that the 
artifacts visible are dominated by the point spread function, rather than calibration errors.

@@

The results of our simulations are shown in Figure~\ref{fig:errors}. The gain errors from our 
EPICal method are shown with solid lines, while the dashed lines represent the visibility-based 
calibration. The thick lines represent the results of using a full sky model, and the thin lines are 
from the single point source model.  In the case of the full sky model, we see the errors trend 
downward with longer integration times, as expected. The EPICal errors are slightly higher than 
the visibility-based errors, on average about 23\% difference. This is not surprising as EPICal 
only uses the information in a single pixel, while the visibilities use all information from the full 
sky model. However, we see that with this perfect model the errors do behave like noise
in the sense that they integrate down with longer integrations with the expected $t^{-1/2}$ 
trend. 
The same level of gain error as visibility based solutions can be achieved with a longer integration time (or larger damping 
factor).

@@

When using the single point source model, all calibration methods in figure~\ref{fig:errors} trend 
downward until they reach an error floor due to the confusion sources that were not modeled. 
The EPICal and visibility-based solutions reach a similar floor, but EPICal achieves a 
marginally lower level (on average 5\% lower). This can be attributed to the same reason 
EPICal underperformed with the full sky model: EPICal only used a single pixel in the sky to 
form its solutions. When using the full sky model, it was down weighting the information in the 
rest of the sky. But when the sky model only contains the bright point source, the single pixel 
used in EPICal is the location where this model is most accurate, effectively down weighting 
the pixels with incomplete sky model. In the regime typical of HERA discussed above, EPICal's 
error floor was 10\% lower than that of visibility-based solutions.

@@

Finally, we found that seven antennas\footnote{LWA antenna IDs 48, 85, 124, 148, 203, 217, 
and 244 were omitted.} produced unstable gain solutions, and in fact corrupted the entire array. 
We therefore omitted these antennas from our analysis, resulting in a total of 248 antennas to 
calibrate. In each calibration loop, we form $C_{i,\hat{\boldsymbol{s}}_0}, i=1,2,\ldots N_a$ and 
update our gain estimates over 10 samples (51.2 ms). We iterate the loop 30 times for a 
total of 1.536 seconds. The results of this calibration experiment are shown in figures~
\ref{fig:data_phase}~--~\ref{fig:data_images}.

@@

The gain amplitudes as a function of calibration iteration are shown in 
figure~\ref{fig:data_amp}. There is a fairly wide range in gain amplitudes (from 0.12 to 0.59). 
However, this is not surprising due to the non-uniformity in the cables from the antennas to the 
receivers. Again, the amplitudes are noisy but relatively flat.

To ensure the level of noise in our solutions is reasonable we compared to the fractional errors
from our simulations shown in Figure~\ref{fig:errors}. We first use the Global Sky Model
\citep{deo08} to generate a mock observed sky at 74.03 MHz using the PyGSM python 
module\footnote{https://github.com/telegraphic/PyGSM} \citep{pri16}. We integrate this model
to estimate a total sky flux power of 171,000~Jy, meaning our calibration pixel containing
Cyg A accounts for  approximately $20,539/171,000 \approx 12\%$ of the total sky power. 
This places us in a regime very similar to the incomplete sky model simulations of the 
previous section, where the calibration pixel accounted for 21.7\% of the sky power. We 
should therefore expect our fractional gain noise levels to be at best around 0.1, the level 
to which the simulations asymptotically approached.

@@

Figure~\ref{fig:data_images} shows the improvement in the images due to our calibration. The 
left panel shows the uncalibrated image integrated over 51.2 ms, 29 kHz. Cyg~A is prominent 
near the center of the image, and Cas~A is also clearly visible in the upper right. The middle 
panel shows the image produced after calibration with identical integration time and bandwidth. 
The sidelobes throughout the image are significantly suppressed, and the galactic plane is 
much more evident, despite only modeling Cyg~A and Cas~A. We also note that the feature 
just to the right of Cyg~A is dimmer in the calibrated image, better matching the expected flux 
from the GSM. 
For reference we show the GSM in the right panel of 
Figure~\ref{fig:data_images}, convolved by the LWA point spread function and weighted
by  two factors of the primary beam to match the holographic frame of the data images.

@@

EPICal can be further improved through several extensions. Here we name a few potential 
considerations for further study. These topics can be thought of as extensions to the white 
``estimate gains" box in figure~\ref{fig:schematic}, which can be performed at a much lower 
cadence than the actual correlations.

>>>>

We show schematically the process of calibrating a direct imaging correlator in 
Fig.~\ref{fig:schematic}. Computationally expensive steps that must be performed in near real-time 
are shown inside the gray box. The uncalibrated antenna signals are tapped out after the 
F-engine and correlated against the output image pixel of interest (i.e., equation~\ref{eq:Cna} is 
computed in near real-time). The correlated values are then used to estimate the gains
using equation~\ref{eq:cal_solution}, and additional fitting if desired. This step does not have
the speed requirement of the steps in the gray box because the correlations have been 
averaged in time.
The gains are then passed back 
to the correlator to update the calibration for subsequent integration intervals.

@@

The calibration loop continues by updating the gain estimates every \caliter~time samples. The 
phases of our gain estimates are shown in Fig.~\ref{fig:sim_phase} for 20 such iterations. 
The phase error plotted is the phase relative to the true gain for each antenna (various colored 
lines). One antenna was used as a reference to fix the absolute phase, so has zero phase 
error. The other 50 antennas are shown to have error spanning $2\pi$ initially, and after about 
10 iterations lock into a solution, settling down to noise levels around iteration 12 (0.12~s). We 
stop the simulation when the updated gains trace the thermal noise of the simulated sources, 
which can be seen by the coherence of the 50 antenna gains after iteration 12.

@@

The estimated gain amplitudes for the simulations are shown in Fig.~\ref{fig:sim_amp}. The 
quantity plotted is the magnitude of the estimated gains over the true gains, $\left|g^{(n)}_a/
g_a\right|$, which places all antennas on the same scale. We can see the amplitudes 
converge toward their true values around the same time as the phases (iteration $\sim$12). At 
the beginning of calibration we can see the importance of the damping factor. At $n=0$, a couple of 
gains are shown to have abnormally high amplitude estimates, notably one about 3.3 times its 
true value (red line). These unbalanced high estimates caused the entire set of gains to be 
under estimated at $n=1$, even with a damping factor of 0.35. By $n=5$ the unbalanced 
amplitudes have been damped out and the calibration continues. Without the damping factor, 
the oscillation seen in the first couple iterations would have been significantly larger and taken 
much longer to fade out.

@@

Images created at the beginning of calibration and at the end are shown in 
Fig.~\ref{fig:sim_images}. Each image is obtained over 400 time sample integration, corresponding to 
all snapshot images created with a given set of gain estimates. The left panel shows the image 
produced with our initialized unity gains. Because the phases are completely random, the 
image is essentially noise with the primary beam evident. After 20 calibration iterations, the image is far 
more clear, shown in the middle panel. 
The ten sources are visible, along with rumble throughout the image. By comparing with
a perfectly calibrated image (right panel), we see that the rumble and the shapes of the
point sources are matched between the two images. We therefore conclude that the 
artifacts visible are dominated by the point spread function, rather than calibration errors.

@@

The results of our simulations are shown in Fig.~\ref{fig:errors}. The gain errors from our 
EPICal method are shown with solid lines, while the dashed lines represent the visibility-based 
calibration. The thick lines represent the results of using a full sky model, and the thin lines are 
from the single point source model.  In the case of the full sky model, we see the errors trend 
downward with longer integration times, as expected. The EPICal errors are slightly higher than 
the visibility-based errors, on average about 23\% difference. This is not surprising as EPICal 
only uses the information in a single pixel, while the visibilities use all information from the full 
sky model. However, we see that with this perfect model the errors do behave like noise
in the sense that they integrate down with longer integrations with the expected $t^{-1/2}$ 
trend. 
The same level of gain error as visibility based solutions can be achieved with a longer integration time (or larger damping 
factor).

@@

When using the single point source model, all calibration methods in Fig.~\ref{fig:errors} trend 
downward until they reach an error floor due to the confusion sources that were not modeled. 
The EPICal and visibility-based solutions reach a similar floor, but EPICal achieves a 
marginally lower level (on average 5\% lower). This can be attributed to the same reason 
EPICal underperformed with the full sky model: EPICal only used a single pixel in the sky to 
form its solutions. When using the full sky model, it was down weighting the information in the 
rest of the sky. But when the sky model only contains the bright point source, the single pixel 
used in EPICal is the location where this model is most accurate, effectively down weighting 
the pixels with incomplete sky model. In the regime typical of HERA discussed above, EPICal's 
error floor was 10\% lower than that of visibility-based solutions.

@@

Finally, we found that seven antennas\footnote{LWA antenna IDs 48, 85, 124, 148, 203, 217, 
and 244 were omitted.} produced unstable gain solutions, and in fact corrupted the entire array. 
We therefore omitted these antennas from our analysis, resulting in a total of 248 antennas to 
calibrate. In each calibration loop, we form $C_{i,\hat{\boldsymbol{s}}_0}, i=1,2,\ldots N_a$ and 
update our gain estimates over 10 samples (51.2 ms). We iterate the loop 30 times for a 
total of 1.536 seconds. The results of this calibration experiment are shown in Figs.~
\ref{fig:data_phase}~--~\ref{fig:data_images}.

@@

The gain amplitudes as a function of calibration iteration are shown in 
Fig.~\ref{fig:data_amp}. There is a fairly wide range in gain amplitudes (from 0.12 to 0.59). 
However, this is not surprising due to the non-uniformity in the cables from the antennas to the 
receivers. Again, the amplitudes are noisy but relatively flat.

To ensure the level of noise in our solutions is reasonable we compared to the fractional errors
from our simulations shown in Fig.~\ref{fig:errors}. We first use the Global Sky Model
\citep{deo08} to generate a mock observed sky at 74.03 MHz using the PyGSM python 
module\footnote{https://github.com/telegraphic/PyGSM} \citep{pri16}. We integrate this model
to estimate a total sky flux power of 171,000~Jy, meaning our calibration pixel containing
Cyg A accounts for  approximately $20,539/171,000 \approx 12\%$ of the total sky power. 
This places us in a regime very similar to the incomplete sky model simulations of the 
previous section, where the calibration pixel accounted for 21.7\% of the sky power. We 
should therefore expect our fractional gain noise levels to be at best around 0.1, the level 
to which the simulations asymptotically approached.

@@

Figure~\ref{fig:data_images} shows the improvement in the images due to our calibration. The 
left panel shows the uncalibrated image integrated over 51.2 ms, 29 kHz. Cyg~A is prominent 
near the center of the image, and Cas~A is also clearly visible in the upper right. The middle 
panel shows the image produced after calibration with identical integration time and bandwidth. 
The sidelobes throughout the image are significantly suppressed, and the galactic plane is 
much more evident, despite only modeling Cyg~A and Cas~A. We also note that the feature 
just to the right of Cyg~A is dimmer in the calibrated image, better matching the expected flux 
from the GSM. 
For reference we show the GSM in the right panel of 
Fig.~\ref{fig:data_images}, convolved by the LWA point spread function and weighted
by  two factors of the primary beam to match the holographic frame of the data images.

@@

EPICal can be further improved through several extensions. Here we name a few potential 
considerations for further study. These topics can be thought of as extensions to the white 
``estimate gains" box in Fig.~\ref{fig:schematic}, which can be performed at a much lower 
cadence than the actual correlations.
